Do music charts still matter in the streaming age? An analysis predicting Beyoncé’s Spotify numbers using a regression model.

### 2024-04-17 paper/paper.qmd
Update written sections in paper

---
title: "Do music charts still matter in the streaming age? An analysis predicting Beyoncé’s Spotify numbers using a regression model."
author: Tam Ly
thanks: "Code and data are available at: [https://github.com/atn-ly/beyonce](https://github.com/atn-ly/beyonce)"
date: today
date-format: long
abstract: "This paper analyzes the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100* utilizing data collected from Spotify and Billboard. We find that the number of Spotify streams of a Beyoncé song increases as its duration on the *Billboard Hot 100* increases. Our discovery highlights the relevance of charts in shaping today’s music industry."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Beyoncé Giselle Knowles-Carter is a singer, songwriter, producer, dancer, and one of the most successful and influential artists of our time. Over the course of her career, she has won 32 Grammy Awards, sold beyond 200 million records, and more than 100 of her songs have landed on the *Billboard Hot 100* chart. The *Billboard Hot 100* is a music ranking system that showcases 100 of the most popular songs in the United States each week. Listeners can stream these hits on Spotify, the world’s most popular audio streaming subscription service.

There is a clear gap in knowing what makes a song famous. In the ever-changing music industry, it is not only about melody and lyrics, but also promotion. In this paper, we focus on music promotion and the importance of Billboard charting. We analyze the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. The estimand is the popularity of a Beyoncé song and this is considered in the number of Spotify streams.

Using data collected from Spotify and Billboard, we found that the number of Spotify streams of a Beyoncé song increases as its duration on the *Billboard Hot 100* increases.

@@

The dataset used in this paper was created using data gathered from Spotify [@spotify] and Billboard [@billboard]. The data was collected by the authors for the purpose of this paper to investigate the number of Spotify streams based on weeks spent on the *Billboard Hot 100* for Beyoncé songs. Song titles, album titles, and the number of Spotify streams were collected from Spotify [@spotify], while weeks spent on the *Billboard Hot 100* were collected from Billboard [@billboard]. There were no similar datasets available that could have been used.

@@

First, we used Spotify [@spotify] to collect data on Beyoncé albums. Here, we had to decide which albums to include in our dataset since Beyoncé has 15 different albums in her Spotify discography. We decided to only include her eight studio albums and one soundtrack album and exclude her live albums and compilation albums because these are not as popular. Furthermore, among her eight studio albums, we only looked at the deluxe versions, since these contain the same songs with the same streaming numbers as the standard version along with some additional tracks.

Next, using Spotify [@spotify], we collected data on Beyoncé songs from the nine albums we selected and had to decide which types of tracks to include. Since we are interested in songs that charted, we decided to exclude remixes, extended mixes, Spanish versions, interludes, and original demos that were included in her albums because these are not as popular. Note that we also did not include songs where Beyoncé was a guest feature on another artist’s song.

Then, we used Spotify [@spotify] to collect data on the number of streams for the songs that we selected. Spotify gives one stream to a track if it is played for at least 30 seconds and updates its streaming numbers once a day at approximately 3 PM EST [@spotify]. We collected this data on April 12, 2024 after 3 PM EST.

Lastly, we collected data on the number of weeks each song spent on the *Billboard Hot 100* from Billboard [@billboard]. Billboard keeps a chart history for Beyoncé that includes the debut date, peak position, peak date, and weeks spent for each song that charted on the *Billboard Hot 100*. From this list, we only collected the number of weeks spent on the chart for our selected songs. Billboard updates the *Billboard Hot 100* as well as the artist’s chart history every Tuesday [@billboard]. We collected this data on Friday, April 12, 2024. For songs that did not chart, we recorded them with a 0 in our dataset.

@@

There are 4 variables in this dataset:

@@

@tbl-variables shows a sample of the dataset with all the variables and the first 10 out of 140 observations.

@@

@fig-variables summarizes the data with all 140 observations. We see a positive correlation between the number of Spotify streams and weeks spent on the *Billboard Hot 100*.

@@

The goal of our modeling strategy is to predict the number of Spotify streams for a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. We used a negative binomial regression model in a Bayesian framework. Negative binomial regression is a type of generalized linear model that is useful for modeling count data.

@@

We expect a positive relationship between the number of Spotify streams and weeks spent on the *Billboard Hot 100* based on the positive correlation in the graph that we observed in @sec-data.

Negative binomial regression operates under several assumptions. It assumes linearity between the outcome and predictor variables, independence of observations, and no multicollinearity.

@@

@@

## First discussion point

## Second discussion point

## Third discussion point

@@

>>>>

---
title: "Do music charts still matter in the streaming age? An analysis predicting Beyoncé’s Spotify numbers using a regression model."
author: Tam Ly
thanks: "Code and data are available at: [https://github.com/atn-ly/beyonce](https://github.com/atn-ly/beyonce)"
date: today
date-format: long
abstract: "Music charts are a powerful marketing tool. This paper analyzes the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100* utilizing data collected from Spotify and Billboard. In our negative binomial regression model, we find that the number of Spotify streams of a Beyoncé song increases as its duration on the *Billboard Hot 100* increases. Our discovery highlights the relevance of charts in shaping today’s music business industry."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Beyoncé Giselle Knowles-Carter is a singer, songwriter, producer, dancer, and one of the most successful and influential artists of our time. Over the course of her career, she has won 32 Grammy Awards, sold beyond 200 million records, and more than 100 of her songs have landed on the *Billboard Hot 100* chart [@sony]. The *Billboard Hot 100* is a music ranking system that is supposed to showcase 100 of the most popular songs in the United States each week [@substack]. Listeners can stream these hits on Spotify, the world’s most popular audio streaming subscription service [@aboutspotify].

There is a clear gap in knowing what makes a song famous. In the ever-changing music industry, it is not only about melody and lyrics, but also promotion. In this paper, we focus on music promotion and the importance of Billboard charting as a marketing tool. We analyze the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. The estimand is the popularity of a Beyoncé song and this is considered in the number of Spotify streams.

We are consequently able to answer the following question: What is the effect of longevity on the Billboard charts on a Beyoncé song’s number of Spotify streams? Using data collected from Spotify and Billboard and a negative binomial regression model, we found that the number of Spotify streams of a Beyoncé song increases as its duration on the *Billboard Hot 100* increases. We deduce that Billboard charting is still important for an artists’ popularity and career.

@@

The dataset used in this paper was created using data gathered from Spotify [@spotify] and Billboard [@billboard]. The data was collected by the authors for the purpose of this paper to investigate the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. Song titles, album titles, and the number of Spotify streams were collected from Spotify, while weeks spent on the *Billboard Hot 100* were collected from Billboard. There were no similar datasets available that could have been used.

@@

First, we used Spotify to collect data on Beyoncé albums. Here, we had to decide which albums to include in our dataset since Beyoncé has 15 different albums in her Spotify discography. We decided to only include her eight studio albums and one soundtrack album and exclude her live albums and compilation albums because these did not chart. Furthermore, among her eight studio albums, we only looked at the deluxe versions, since these contain the same songs with the same streaming numbers as the standard version along with some additional tracks.

Next, using Spotify, we collected data on Beyoncé songs from the nine albums we selected and had to decide which types of tracks to include. Since we are interested in songs that charted, we decided to exclude remixes, extended mixes, Spanish versions, interludes, and original demos that were included in her albums because these are not as popular. Note that we also did not include songs where Beyoncé was a guest feature on another artist’s song.

Then, we used Spotify to collect data on the number of streams for the songs that we selected. Spotify gives one stream to a track if it is played for at least 30 seconds and updates its streaming numbers once a day at approximately 3 PM EST. We collected this data on April 12, 2024 after 3 PM EST.

Lastly, we collected data on the number of weeks each song spent on the *Billboard Hot 100* from Billboard. Billboard keeps a chart history for Beyoncé that includes the debut date, peak position, peak date, and weeks spent for each song that charted on the *Billboard Hot 100*. From this list, we only collected the number of weeks spent on the chart for our selected songs. Billboard updates the *Billboard Hot 100* as well as the artist’s chart history every Tuesday. We collected this data on Friday, April 12, 2024. For songs that did not chart, we recorded them with a 0 in our dataset.

@@

There are 4 variables in the dataset that we created:

@@

To visualize our data, we create a table and graph. @tbl-variables shows a sample of the dataset with all the variables and the first 10 out of 140 observations. Furthermore, @fig-variables show a scatterplot of the data with all 140 observations. We see a positive correlation between the number of Spotify streams and weeks spent on the *Billboard Hot 100*.

@@

@@

The goal of our modeling strategy is to predict the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. We used a negative binomial regression model in a Bayesian framework. Negative binomial regression is a type of generalized linear model that is useful for modeling count data.

@@

We expect a positive relationship between the number of Spotify streams and weeks spent on the *Billboard Hot 100* based on the positive correlation in the scatterplot graph that we observed in @sec-data.

Negative binomial regression operates under several assumptions. It assumes linearity between the outcome and predictor variables, independence of observations, and no multicollinearity [@negativebinomial].

@@

@tbl-results demonstrates what was expected: the positive number 0.094 means that there is a positive relationship between the number of Spotify streams of a Beyoncé song and the number of weeks it spent on the *Billboard Hot 100*. @fig-results further illustrates this and makes it clear that we expect a positive relationship.

@@

## Findings

In this paper, we collected and visualized data from Spotify and Billboard as discussed in @sec-data. Then, we used a Bayesian negative binomial regression model as seen in @sec-model to predict the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. Following our analysis in @sec-results, we found that the number of Spotify streams of a Beyoncé song increased as its duration on the *Billboard Hot 100* increased. We will now discuss the historical context behind Billboard and why huge superstars, such as Beyoncé, still care about charting.

## A bit of history

In the 1940s, Billboard started developing music charts to try and determine the most popular song [@substack]. This was to help inform people in the music industry, including record stores and jukebox operators, about which records to order [@substack]. The *Billboard Hot 100* was created in 1958 and has undergone complete upheavals since its inception [@substack]. How exactly does Billboard determine the most popular song? This has always been the subject of much controversy due to its biased survey methods. Billboard first measured sales by surveying record stores, although it was impossible to know if the store clerks were lying or not [@substack]. Then, co-founder of Casablanca Records, Larry Harris, alleged that industry insiders had heavy influence over the charts [@substack]. In the 1990s, Billboard began tracking sales based on barcode scans and hip-hop sales seemingly rose dramatically overnight [@substack].

Charts are supposed to reflect the industry’s reality and in 2018, it was announced that the *Billboard Hot 100* would take streaming data into account [@substack]. However, the exact formula behind Billboard’s calculations remains unknown to the public. We only know that one stream does not necessarily equal one sale and that some streams, such as for on demand services and from paid-tier users, count more than others [@substack]. By the 2010s, the question of what the Billboard charts actually reflect became even more prevalent after fans proved that it was susceptible to coordinated manipulation [@substack].

## Why charts still matter

Do artists as big as Beyoncé still care about music charts? Chart success can bring significant music promotion to an artist’s career. Songs that land on the *Billboard Hot 100* also land on Spotify’s biggest playlists, such as Today’s Top Hits which has almost 35 million followers [@todaystophits]. Cherie Hu, a music industry analyst at the intersection of music and data, has deemed playlists “the new album” and “more powerful than radio” as a placement on Today’s Top Hits can bring up to $163,000 in additional revenue per track [@cherie]. Moreover, for new artists, high positions in the charts can lead to more ownership and creative control in their musical career. This can help explain the importance of why many of Beyoncé’s longest charting songs come from her earlier albums. For more established artists who have a loyal fanbase, charting is essential for maintaining their brand and curtailing any image of a decline.

@@

Some weaknesses become apparent in our approach in using Spotify streams as a metric for the popularity of a Beyoncé song. First, one of Beyoncé’s albums, *Lemonade*, was added to Spotify three years after its release. This means that the number of Spotify streams for the songs from this album are underrepresented in our dataset. Furthermore, Spotify numbers are also not immune to industry politics that underlie promotional mechanics. All of the major record labels own a sizable share of Spotify and can give some artists preferential treatment on the platform. For example, Beyoncé is under Sony Music which owns a 2.65% stake in Spotify [@owners]. Lastly, we did not consider that Beyoncé’s older songs have had more time to accumulate streams than her latest work. For instance, *COWBOY CARTER* was just released on March 29, 2024 while *Dangerously in Love* was released on June 20, 2003 [@spotify].

In the future, it could be on interest to run the analysis again to give Beyoncé’s latest album *COWBOY CARTER* more time to see how long it lasts on the chart. We could also standardize the length to accumulate streams such as one year after its release for consistency. Another possible idea for future work could be to use a mix of new and established artists to see if we get the same result.
